---
title: "T test and Box plot and Life Expectancy Gap"
author: "Riley"
format: pdf
editor: visual
---

```{r}
# Load necessary libraries
library(dplyr)
library(ggplot2)
library(ggrepel)  # install if necessary
```

```{r}
# Load CSV
df <- read.csv("/Users/rileynewton/Desktop/cleandata.csv")  # update path

# Remove unwanted columns
df <- df %>% select(-c(X, X.1, X.2, X.3, X.4, X.5, X.6))

# Create life expectancy gap
df$LifeExpectancyGap <- df$`Female.Life.Expectancy..years.` - df$`Male.Life.Expectancy..years.`
```

```{r}
# Paired t-test: do women live longer than men?
t_test_result <- t.test(
  df$`Female.Life.Expectancy..years.`,
  df$`Male.Life.Expectancy..years.`,
  paired = TRUE,
  alternative = "greater"
)
t_test_result
```

```{r}
# Boxplot of male vs female life expectancy
df_long <- df %>%
  select(Area, `Male.Life.Expectancy..years.`, `Female.Life.Expectancy..years.`) %>%
  tidyr::pivot_longer(cols = c(`Male.Life.Expectancy..years.`, `Female.Life.Expectancy..years.`),
                      names_to = "Gender", values_to = "LifeExpectancy") %>%
  mutate(Gender = ifelse(Gender == "Male.Life.Expectancy..years.", "Male", "Female"))

ggplot(df_long, aes(x = Gender, y = LifeExpectancy, fill = Gender)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.5) +
  labs(title = "Life Expectancy by Gender Across U.S. States",
       y = "Life Expectancy (years)",
       x = "") +
  theme_minimal() +
  scale_fill_manual(values = c("Male" = "steelblue", "Female" = "pink"))
```

A paired-samples t-test showed that women have a significantly higher life expectancy than men across U.S. states, **t**(51) = 68.01, **p** \< .001. On average, women live **5.68 years longer** than men (95% CI: 5.54, ∞).

Finding if Female-Male Age gaps is larger in states with lower total life expectancy:

```{r}
# Advanced life expectancy gap plot
df <- df %>%
  mutate(LifeExpCategory = ifelse(`Total.Life.Expectancy..years.` < 78, "Low", "High"))

gap_plot <- ggplot(df, aes(x = `Total.Life.Expectancy..years.`, y = LifeExpectancyGap, color = LifeExpCategory)) +
  geom_point(size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, color = "blue", linetype = "dashed") +
  geom_text_repel(aes(label = Area), size = 3, max.overlaps = 15) +
  scale_color_manual(values = c("Low" = "red", "High" = "green")) +
  labs(
    title = "Gender Life Expectancy Gap vs Total Life Expectancy by State",
    subtitle = "States with lower total life expectancy tend to have higher female-male gaps",
    x = "Total Life Expectancy (years)",
    y = "Female − Male Life Expectancy Gap (years)",
    color = "Total Life Expectancy Category"
  ) +
  theme_minimal(base_size = 12)

print(gap_plot)
```

-   cor = -0.577 This is a moderate-to-strong negative correlation.

    -   Interpretation: as total life expectancy decreases, the female–male gap tends to increase. In other words, states where people live shorter lives tend to have a bigger difference between women’s and men’s life expectancy.

-   t = -4.99, df = 50, p-value = 7.688e-06

    -   The p-value is extremely small (\< 0.001).

    -   This means the correlation is statistically significant — it’s very unlikely this negative relationship is due to chance.

-   -0.734 to -0.360

    -   This tells you that, with 95% confidence, the true correlation in the population is between -0.73 and -0.36.

    -   Since the entire interval is negative, it confirms the negative relationship.

```{r}
# Linear regression: LifeExpectancyGap ~ Total.Life.Expectancy
lm_model <- lm(LifeExpectancyGap ~ `Total.Life.Expectancy..years.`, data = df)
summary(lm_model)
```
